Parsing Postfix log files

### 2008-07-07 doc/logparser-background.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

Ten other parsers have been reviewed in
appendix~\refwithpage{other-parsers} as part of the background research for
this project.  None of the reviewed parsers perform the type of advanced
parsing and log correlation described here; all are intended to perform a
specific parsing and reporting task, rather than be a generic parser,
extracting data and leaving generation of reports from the data to other
programs.  Some parsers save the data extracted to a data store, but the
many discard all data once they have finished running, making historical
analysis impossible.  The other parsers reviewed all produce a report of
varying complexity and detail, whereas the program described here doesn't
attempt to produce a report at all; that responsibility is deferred to a
separate program, to be developed later.  The parsing algorithm and program
described here are designed to enable much more detailed log analysis by
providing a stable platform for subsequent programs to develop upon.

>>>>

This section provides background information helpful in understanding the
remainder of the document.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
document, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this document.  Other projects which attempt to
parse Postfix log files are summarised (full details are available in
\sectionref{other-parsers}), finishing with a review of previously
published research in this area.

@@

Ten other parsers have been reviewed in \sectionref{other-parsers} as part
of the background research for this project.  None of the reviewed parsers
perform the type of advanced parsing and log correlation described here;
all are intended to perform a specific parsing and reporting task, rather
than be a generic parser, extracting data and leaving generation of reports
from the data to other programs.  Some parsers save the data extracted to a
data store, but the many discard all data once they have finished running,
making historical analysis impossible.  The other parsers reviewed all
produce a report of varying complexity and detail, whereas the program
described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

### 2008-07-07 doc/logparser-database-schema.tex
Changed RBL to DNSBL, and added a reference for it.
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

    \item [start] The timestamp of the first log line, in seconds since the
        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an \RBL{}\@.

>>>>

    \item [start] The timestamp of the first log line, in seconds since the
        epoch (explained in the glossary, \sectionref{Glossary}).

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an \DNSBL{}\@.

### 2008-07-07 doc/logparser-introduction.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check whether new anti-spam measures are working.  The
more adventurous will have generated statistics to show how many hits each
of their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\RBL{}\footnote{This document is supplied with a glossary, see
appendix~\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different versions of Postfix.  This paper
documents the difficult process of parsing Postfix log files, presenting
\PLP{}, a program which parses log files and places the resulting data into
a database for later use.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are considered, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, utilising the graphs in
appendix~\refwithpage{graphs} for illustrative purposes.  The section
concludes with a description of using the tools provided with the parser to
generate new rules (specifically the \regex{} in each rule) from unparsed
log lines.

>>>>

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check whether new anti-spam measures are working.  The
more adventurous will have generated statistics to show how many hits each
of their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the author's first real foray into processing
Postfix log files.}  Very few will have performed in-depth parsing and
analysis of their log files, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix log files, where each log line was added on
an ad hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\DNSBL{}\footnote{This document is supplied with a glossary, see
\textsection\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different versions of Postfix.  This paper
documents the difficult process of parsing Postfix log files, presenting
\PLP{}, a program which parses log files and places the resulting data into
a database for later use.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately; identifying regular high volume uses (e.g.\
customer newsletters) and restricting those uses to off-peak times;
detecting virus outbreaks which propagate via email; as a base for billing
customers on a shared server.  Preserving the raw data enables users to
develop a multitude of uses far beyond those conceived of by the author.

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are considered, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, utilising the graphs in
\sectionref{graphs} for illustrative purposes.  The section concludes with
a description of using the tools provided with the parser to generate new
rules (specifically the \regex{} in each rule) from unparsed log lines.

### 2008-07-07 doc/logparser-parsing-rules.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

    \item The topic of rule efficiency was discussed next, covering the
        effects of caching compiled \regexes{} and optimal ordering of
        rules, with reference to the graphs in
        appendix~\refwithpage{graphs}.

>>>>

    \item The topic of rule efficiency was discussed next, covering the
        effects of caching compiled \regexes{} and optimal ordering of
        rules, with reference to the graphs in \sectionref{graphs}.
